---
title: 'Analysis of the Effect of SARS-CoV-2 on Human and Non-Human Tissues'
author: "Daniela Dueñas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
csl: bioinformatics.csl
bibliography: references.bib
params:
  gff_file_path: "data/metadata/sars_refgenome_annotation.gff"
  vcf_dir_path: "data/11_vcf_output_for_R"
  sra_runtable_path: "data/00_sra_runtable/2024-12-01_clean_data.csv"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE)
```

# Background and Overview

This is a report on SARS-CoV-2, including some variant analysis [@koyama2020variant].

# Methods

See the set of tutorials on the [vcfR package website](https://knausb.github.io/vcfR_documentation/index.html).

You may also want to use any of a range of different COVID data packages and data sources:

* https://kjhealy.github.io/covdata/
* https://github.com/como-ph/oxcovid19
* https://ropensci.org/blog/2020/10/20/searching-medrxivr-and-biorxiv-preprint-data/
* https://covidtracking.com/data/api
  * `readr::read_csv("https://api.covidtracking.com/v1/states/daily.csv")`

## Subsections are ok too

# Results and Discussion

```{r load-packages-and-functions}
library("vcfR")
library("ggplot2")
library("dplyr")
library("ggthemes")

# this loads all function files in the code/functions directory
function_paths <- list.files("code/functions", full.names = TRUE)
purrr::walk(function_paths, source)
```

```{r load-vcf-data}
# load in, tidy, and stack all of the VCF files in the target directory
# these are the output of the bash pipeline
stacked_vcfs <- parse_tidy_and_stack_vcfs(vcf_dir_path = params$vcf_dir_path)

# load in the gff file with genome annotations for SARS-CoV-2
gff <- read_gff(gff_file_path = params$gff_file_path)

# pull out just the named genes from the annotation file
gene_table <- extract_genes_from_gff(annotation_object = gff)

# combine the stacked vcf table with the gene names and the metadata from
# the sra runtable file
vcf_with_metadata <-
  add_genes_metadata_to_vcfstack(sra_runtable_path = params$sra_runtable_path,
                                 stacked_vcf = stacked_vcfs,
                                 cleaned_genes_table = gene_table)
```

# Figures

```{r distinct-snp-plot}
# create a plot of unique SNP locations within each gene across all samples
vcf_with_metadata %>%
  filter(!is.na(gene)) %>% # get rid of SNPs not in gene regions
  group_by(gene, pos) %>%
  tally() %>% # this gives a column n for the number of gene by position
  group_by(gene) %>%
  tally() %>% # this collapses that down to the number of unique SNP locations
  ggplot(aes(x = gene,
             y = n)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Count of distinct SNPs in Named SARS-CoV-2 Genes",
       x = "Gene Name") +
  theme_few() # get rid of the grey background
ggsave("output/distinct-snp-plot.png", width = 8, height = 6)
```

```{r sample-number-plot}
# A plot of the number of sample runs per organism
vcf_with_metadata %>%
  filter(!is.na(gene)) %>% # get rid of SNPs not in gene regions
  group_by(sample, organism) %>%
  tally() %>% # this gives a column n for the number of samples per organism
  group_by(organism) %>%
  tally() %>%
  ggplot(aes(x = organism,
             y = n)) +
  geom_col(fill = "blue") +
  geom_text(aes(label = n), vjust = -0.1, color = "black") +
  labs(title = "Number of Samples Run Per Organism Type",
       x = "Organism") +
  theme_few()
ggsave("output/sample-number-plot.png", width = 8, height = 6)
```

```{r qual-scatterplot}
# A scatterplot of SNP quality scores in human and C. aethiops genomes
vcf_with_metadata %>%
  filter(qual > 200) %>%
  mutate(qual = as.numeric(qual), # converts qual to numeric values
         qual = round(qual)) %>% # rounds to nearest whole number
  ggplot(aes(x = pos,
             y = qual,
             color = organism)) +
  geom_point(alpha = 0.6) +
  labs(title = "SNP Quality Scores Across Organism Genomes",
       x = "Position on Genome",
       y = "SNP Quality",
       color = "Organism") +
  theme_few()
ggsave("output/qual-scores_scatterplot.png", width = 8, height = 6)
```

```{r human-tissue-plot}
# A scatterplot of SNP quality scores from various human tissues
vcf_with_metadata %>%
  filter(organism == "Homo sapiens") %>% # filters samples to only H. sapiens
  filter(qual > 200) %>%
  mutate(qual = as.numeric(qual), # converts qual to numeric values
         qual = round(qual)) %>% # rounds qual scores to nearest whole number
  ggplot(aes(x = pos,
             y = qual,
             color = tissue)) +
  geom_point(alpha = 0.6) +
  labs(title = "SNP Quality Scores Across Human Tissue Samples",
       x = "Position on Genome",
       y = "SNP Quality",
       color = "Tissue Type") +
  theme_few()
ggsave("output/human-tissue-qual_scatterplot.png", width = 8, height = 6)
```

```{r model-tissue-plot}
# A scatterplot of SNP quality scores from C. aethiops tissue (kidney)
vcf_with_metadata %>%
  filter(organism == "Chlorocebus aethiops") %>% # filters out H. sapiens
  filter(qual > 200) %>%
  mutate(qual = as.numeric(qual), # converts qual to numeric values
         qual = round(qual)) %>% # rounds qual scores to nearest whole number
  ggplot(aes(x = pos,
             y = qual,
             color = tissue)) +
  geom_point(alpha = 0.6) +
  labs(title = "SNP Quality Scores Across Chlorocebus aethiops Tissue Samples",
       x = "Position on Genome",
       y = "SNP Quality",
       color = "Tissue Type") +
  theme_few()
ggsave("output/model-tissue-qual_scatterplot.png", width = 8, height = 6)
```

```{r treatment-human-plot}
# Scatterplot of SNP quality scores of mock/SARS infected cells in H. sapiens
vcf_with_metadata %>%
  filter(organism == "Homo sapiens") %>%
  filter(qual > 200) %>%
  mutate(qual = as.numeric(qual), # converts qual to numeric values
         qual = round(qual)) %>% # rounds qual scores to nearest whole number
  ggplot(aes(x = pos,
             y = qual,
             color = treatment)) +
  geom_point(alpha = 0.6) +
  labs(title = "SNPs Across Human Genome Based on Cell Treatment",
       x = "Position on Genome",
       y = "SNP Quality",
       color = "Treatment Type") +
  theme_few()
ggsave("output/human-treatments_scatterplot.png", width = 8, height = 6)
```

```{r treatment-model-plot}
# Scatterplot of SNP quality scores of mock/SARS infected cells in C. aethiops
vcf_with_metadata %>%
  filter(organism == "Chlorocebus aethiops") %>%
  mutate(qual = as.numeric(qual), # converts qual to numeric values
         qual = round(qual)) %>% # rounds qual scores to nearest whole number
  ggplot(aes(x = pos,
             y = qual,
             color = treatment)) +
  geom_point(alpha = 0.6) +
  labs(title = "SNPs Across C. aethiops Genome Based on Cell Treatment",
       x = "Position on Genome",
       y = "SNP Quality",
       color = "Treatment Type") +
  theme_few()
ggsave("output/model-treatments_scatterplot.png", width = 8, height = 6)
```

**Figure 1**: N and S genes have more unique SNPs in the set of samples analyzed.

# Tables

```{r example-table}
# An example table to show the length of each gene using its start and end
gene_table <- gene_table %>%
  mutate(length = end - start) %>%
  select(gene_name, start, end, length) %>%
  knitr::kable(col.names = c("Gene Name",
                             "Start",
                             "End",
                             "Length"))
write.csv(gene_table, "output/gene_table.csv", row.names = FALSE)
```

**Table 1**: Gene names, locations, and lengths in the SARS-CoV-2 genome. Higher SNP counts in the S and N genes may be related to the larger size of these genes.

```{r SNP-tally}
# A table of SNPs with quality values greater than 200 from each gene
snp_tally <- vcf_with_metadata %>%
  filter(qual > 200) %>%
  group_by(gene) %>%
  tally() %>%
  knitr::kable(col.names = c("Gene Name",
                             "Number of SNPs"))
write.csv(snp_tally, "output/snp_tally.csv", row.names = FALSE)
```

```{r top-snp-tally}
# A table including information about SNPs with qual > 200
snp_tally_data <- vcf_with_metadata %>%
  filter(qual > 200) %>%
  group_by(gene, organism, tissue, treatment) %>%
  tally() %>%
  rename("Gene Name" = gene,
         "Organism" = organism,
         "Tissue" = tissue,
         "Treatment" = treatment,
         "Number of SNPs" = n) %>%
  arrange(desc(`Number of SNPs`)) %>% # arranges values in descending order
  knitr::kable()
write.csv(snp_tally_data, "output/snp_tally_data.csv", row.names = FALSE)
```

# Sources Cited
